---
title: "Phyloseq PreProcessing"
author: "Calla Bush St George"
date: "`r Sys.Date()`"
output: html_document
  toc: yes
  toc_float: 
    collapsed: no
    smooth_scroll: yes
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/02_PreProcessing/")
# Send any figure output to the 02_PreProcessing folder
```


# Load libraries
```{r load-libraries}
library("devtools")
#install_github("joey711/phyloseq")
library("phyloseq")
library("tidyverse")
```

# Goals

Here, we will process the data into a phyloseq object.

- ASV table
- Taxonomy table
- Track Reads (metadata)

Then, we will remove the following:

1. Remove Chloroplasts.
2. Remove Mitochondria.
3. Remove samples without "enough" reads.

Finally, write a data file of phyloseq output. 

# Load data

## ASV table
```{r load-ASV-table}
# First, load ASV table
load("data/01_DADA2/ASV_counts.RData")

# Inspect ASV_tab
head(asv_tab)[,1:5]

# Fix names
sample_names <- colnames(asv_tab)
samples_fixed <- sapply(strsplit(basename(sample_names), "_"), `[`,1)
head(samples_fixed)

# re-write the ASV count file to fix names
colnames(asv_tab) <- samples_fixed
str(asv_tab)
```

## Taxonomy table
```{r load-tax-table}
tax_df <- read.table("data/01_DADA2/ASV_taxonomy.tsv", sep = "\t", skip = 1)
head(tax_df)

# fix column names
colnames(tax_df) <- c("asv_names", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "ASV", "ASVseq")

head(tax_df)

#Taxonomy Table Matrix
tax_mat <-
tax_df %>%
  tibble::column_to_rownames(., var = "asv_names") %>%
  as.matrix()
```

## Track Reads Data
```{r load-track-reads}
load("data/01_DADA2/track_read_counts.RData")

# Take a look at the data
track_counts_df

# Make sure all data in ASV table matches ASV taxonomy
head(track_counts_df)
dim(track_counts_df)

# Please in the terminal copy metadata.csv to data folder 
# into: /workdir/cab565/git_repos/SalinityGradient_16S/data
# from: /workdir/in_class_data/SalinityGradient_16S/metadata.csv

#Load in metadata
metadata_df <- read.csv("data/metadata.csv")
head(metadata_df)
dim(metadata_df)

# Merge metadata df with track_reads_df
head(metadata_df)
head(track_counts_df)

metadata_track_reads_df <-
  metadata_df %>% 
  left_join(., track_counts_df, by = "names") %>%
  dplyr::select(-X)

head(metadata_track_reads_df)

# Update row.names to be sample names
## Before
row.names(metadata_track_reads_df)
## Rewrite
row.names(metadata_track_reads_df) <- metadata_track_reads_df$names
## Check that it worked
row.names(metadata_track_reads_df)
# Intuition check
head(metadata_track_reads_df)
```

# Handoff to phyloseq
```{r phyloseq-handoff}
# Double check it's all good
dim(asv_tab)
dim(tax_mat)

# Intuition check
stopifnot(row.names(asv_tab) == row.names(tax_mat))

# Construct the phyloseq object
raw_physeq <- phyloseq(otu_table(asv_tab, taxa_are_rows = TRUE), 
                       sample_data(metadata_track_reads_df),
                       tax_table(tax_mat))
raw_physeq

# Save this raw phyloseq object
save(raw_physeq, file = "data/02_PreProcessing/raw_physeq.RData")
```

# Clean up the data

Remove: 

1. Chloroplasts
2. Mitochondria

```{r rm-mitos-chloros}
#Remind myself of the tax matrix
#View(tax_mat)

# Make a new physeq without chloroplasts
noChloros_physeq <-
  raw_physeq %>%
  # rm chloroplasts
  subset_taxa(Order != "Chloroplast" | is.na(Order))

# How many taxa were chloroplasts
num_chloro_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
num_chloro_ASVs

#Intuition check
#noChloros_physeq %>%
#  tax_table() %>%
#  data.frame() %>%
#  View()

#Remove mitochondria
noChlorosMitos_physeq <-
  noChloros_physeq %>%
  # rm chloroplasts
  subset_taxa(Family != "Mitochondria" | is.na(Family))

# How many mitochondrial ASV?

num_mito_ASVs <- ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)
num_mito_ASVs

noChlorosMitos_physeq

# How many total ASVs were removed from Chloros and Mitos?
ntaxa(raw_physeq) - ntaxa(noChlorosMitos_physeq)
# Proportion of ASVs?
ntaxa(noChlorosMitos_physeq)/ntaxa(raw_physeq)
```

# Evaluate and remove the control samples

Take a look at the negative controls and then make a decision about whether or not to remove the ASVs that we found in our controls

1. Negative controls
2. ASVs found within the negative controls and their distribution in the samples. 
3. Evaluate the mock community

## Negative Controls
```{r neg-controls}
# Create a vector of samples that are negative controls
control_samples <- c("WaterControl", "022um-Control", 
                     "3um-Control", "DNA-Ext-Control")

# Make a new phyloseq object of only the control samples
control_physeq <-
  noChlorosMitos_physeq %>%
  subset_samples(., names %in% control_samples) %>%
  # Force remove ASVs with 0 counts
  prune_taxa(taxa_sums(.) > 0, .)

control_physeq

# Vector of control ASVs
control_ASVs <-
  control_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(ASV) %>%
  as.vector()

# Visualize raw abundance values of control ASVs
control_physeq %>%
  plot_bar(., "ASV", fill = "ASV")

################# Check the control ASVs in the samples
# Make a new phyloseq object with only the control ASVs and all samples
control_ASV_physeq <-
  noChlorosMitos_physeq %>%
  subset_taxa(., ASV %in% control_ASVs$ASV)

# Inspect it
control_ASV_physeq

# Let's take a look at the abundances
control_ASV_physeq %>%
  plot_bar(., "ASV", fill = "ASV", facet_grid = Sample_or_Control~.)

#Check tax of ASV 3
#control_ASV_physeq %>% 
#  tax_table() %>%
#  data.frame() %>%
#  View()

# Would only like to keep ASV 18 and ASV 56. Everything else is trash.
control_ASVs_toREMOVE <- dplyr::setdiff(control_ASVs$ASV,
                                        c("ASV_18", "ASV_56"))

# Intution check 
control_ASVs_toREMOVE
length(control_ASVs_toREMOVE)

### Make new phyloseq object
noControlsChlorosMitos_physeq <-
  noChlorosMitos_physeq %>%
  # Actually remove the control ASVs except 18 and 56
  subset_taxa(., !(ASV %in% control_ASVs_toREMOVE)) %>%
  # Remove control samples
  subset_samples(., !(names %in% control_samples))

# Inspect no controls object
noControlsChlorosMitos_physeq

# After filtering, how many ASVs were removed?
total_ASVs_rm <- ntaxa(raw_physeq) - ntaxa(noControlsChlorosMitos_physeq)

```

##Session information
```{r session-info}
#Ensure reproducibility
devtools::session_info()
```




